---
title: "chap.1 pattern detection thresholds"
author: "Arthur de Grandpré"
date: "26 février 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Context

This is an initial draft to the first chapter of my thesis about the resolution threshold of vegetation pattern detection by remote sensing tools.  
The main idea of this chapter is to:  

Compute multiple *indicators* describing a pattern for various known patterns in a very high resolution remote sensing image, and then downsample the resolution and recompute the indicators.

# Test run

First, a test will be conducted on a single site.  
It is based on a resampled 30cm image at 1m / pixel and will be used to obtain patch and class metrics of vegetation cover for a zone of about 150m2. 
  
```{r libraries, warning=FALSE, message=FALSE}

library(sp)
library(raster)
library(rgdal)
library(SDMTools)

```
  
```{r test site}

r1=brick("data/ex1/test_CIR_clip_01.tif", crs="+init:epsg=26918") #small subset of 2012 mosaic made in arcmap with crs nad83 utm 18n
r1=r1/255 # this is to bring back the reflectance values to a 0-1 scale
r1$ndvi=(r1$test_CIR_clip_01.1-r1$test_CIR_clip_01.2)/(r1$test_CIR_clip_01.1+r1$test_CIR_clip_01.2) # this computes the normalized difference vegetation index
plot(r1) # we see the 3 bands and the ndvi index
hist(r1) # we see the value distribution of every band
plotRGB(r1, scale=1, stretch="lin") # this is as CIR imagery (NIR - Red - Green)

```

This is the initial test rasterbrick that will be used for this exercise. it contains 4 bands: NIR, RED, GREEN and NDVI

## landscape metrics

```{r mask}

r2=r1 # to work on a different raster than the original
r2[r2$ndvi<(-0.45) | r2$test_CIR_clip_01.3>0.42]=-10 # a specific mask (won't be easy to generalize)
plot(r2) # every bands were subsetted by the mask
plotRGB(r2, scale=1, stretch="lin") #this is CIR imagery (NIR Red Green)

```

This new raster will be converted to binary values in the next step

```{r binarize}

r3=r2$ndvi #since we only need a single band
r3[r3>(-10)]=1 ; r3[r3==(-10)]=0 # everything assigned to -10 is now 0, and everything else is now 1
plot(r3)

```

This is the binary raster of vegetation cover. From this raster, let's create polygons in order to obtain our patch metrics

```{r polygonize}

p=rasterToPolygons(r3, dissolve=T) # very slow, should look into python & polygonizeR
p=disaggregate(p)

```

```{r filter patch area stats}

pr=rasterize(p,r3) # this creates a raster of p, based on the configuration of r3
plot(pr) # pr contains a raster of 774 patches of different codes

ps=PatchStat(pr) # if we obtain PatchStats from this raster, it creates a statistic for every single patches, including single pixels and including non vegetation

p@data$patchID=rownames(p@data) # adds a column in p that contains the rownames (polygonID)
p2=merge(p,ps, by="patchID") # adds the statistics to the polygons so they contain their information

p2=p2[complete.cases(p2@data),] # removes polygons that do not have all information (single pixels)
rc=rasterize(p2, r3, field="ndvi") # forms a new raster without the single pixels

pv=p2[p2@data$ndvi==1,] # keeps only vegetation pixels
pr2=rasterize(pv, r3) # creates a new raster where only vegetated patches with more than 1 pixel exist
plot(pr2)

```

```{r Class and Patch stats}

PS = PatchStat(pr2) # calculates statistics for every patches
CS = ClassStat(rc) # calculates statistics for the whole vegetation class

```
